Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-05-01 E-Hypergraphs-LICS24.tex
halfway through amended definitions of \consistency and <

\begin{theorem}[Existence of pushouts in $\catname{EHyp_{\Sigma}}$]
\label{th:existence_of_pushouts}
Consider the following span in $\catname{EHyp_{\Sigma}}$
% https://q.uiver.app/#q=WzAsMyxbMCwwLCJaIl0sWzIsMCwiWCJdLFswLDIsIlkiXSxbMCwxLCJmIl0sWzAsMiwiZyIsMl1d
\[\begin{tikzcd}
	Z && X \\
	\\
	Y
	\arrow["f", from=1-1, to=1-3]
	\arrow["g"', from=1-1, to=3-1]
\end{tikzcd}\]
such that
\begin{enumerate}
\label{pushout:assumptions}
    \item $Z$ is a \textit{discrete} e-hypergraph.
    \item $f,g$ are monos.
    \item $[f_{V}(v_i)) = [f_{V}(v_j))$ and $[g_{V}(v_i)) = [g_{V}(v_j))$ for all $v_{i},v_{j}$ in $V_{Z}$.
    \item If $[f_{V}(v)) \not = \varnothing$ then $[g_{V}(v)) = \varnothing$ and if $[g_{V}(v)) \not = \varnothing$ then $[f_{V}(v)) = \varnothing$.
    \item $\consistency(f_{V}(v_i)) = \consistency(f_{V}(v_j))$ and $\consistency(g_{V}(v_i)) = \consistency(g_{V}(v_j))$ for all $v_i,v_j$ in $V_{Z}$.
\end{enumerate}    
then the pushout $X +_{f,g} Y$ exists.
\end{theorem}
\begin{proof}
    Consider the diagram below.
    \[
        \adjustbox{scale=1.25}{
            \begin{tikzcd}
            Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
            Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                            &                                                                  & Q
            \end{tikzcd}}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, <_{X+Y} \}
    \]
    where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, <_{X+Y}, \consistency_{X+Y}$.
    We will omit labels as they are irrelevant to pushout construction.
    Consider relations
    \begin{align*}    
    S_{V} &= \{\\
          &(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}
        \\
        \\
    S_{E} &= \{(x,x) \text{ where } x \in E_{X} + E_{Y}\}
    \end{align*}
    % \begin{align*}
    % S_{E} &= \{\\
    % &(x_i,y_j) \in (E_{X} + E_{Y}) \times (E_{X} + E_{Y})\; |\\
    % &\exists z \in E_{Z} \; . \; x_i = f_{E};\iota_{1,E}(z) \text{ and } y_j = g_{E};\iota_{2,E}(z)\\
    % &\text{ where $x_i \in E_{X}$ and $y_j \in E_{Y}$}\\
    % &\}
    % \end{align*}
and let relations $R_{V},R_{E}$ be their reflexive symmetric and transitive closures respectively.

We then quotient the set vertices and edges in $X + Y$ by $R_{V}$ and $R_{E}$ respectively.
    \begin{align*}
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = \{&\\
            &\sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})}, \sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})}, \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})}, \consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}}, <_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\\
        & \}    
    \end{align*}
We will refer to $\sim (R_{V},R_{E})$ just as $\sim$, \textit{i.e.}, by writing $\sfrac{V_{X} + V_{Y}}{\sim}$ and concrete relation will be clear from the context.
We have 
\[
    \sfrac{s_{X+Y}}{\sim} : \sfrac{E_{X} + E_{Y}}{\sim} \to (\sfrac{V_{X} + V_{Y}}{\sim})^{*}
\]
and there is an obvious surjective function $[-]_{V} : (V_{X} + V_{Y}) \to (\sfrac{V_{X} + V_{Y}}{\sim})$ that maps elements to their equivalence classes and $[-]_{V}^{*}$ is its extension to sequences.
There is also $[-] : E_{X} + E_{Y} \to \sfrac{E_{X} + E_{Y}}{\sim}$ and we will omit subscripts as the correct type will be clear from the argument.
We then define 
\[
    \sfrac{s_{X+Y}}{\sim}([e]) = s_{X+Y};[-]^{*}(e) = [s_{X+Y}(e)]^{*}
\]
We will also use subscripts when it is important to tell if an element of $E_{X} + E_{Y}$ has a pre-image in either $E_{X}$ or $E_{Y}$ by writing $e_{x}$ or $e_{y}$.
Likewise, 
\[
    t_{\sfrac{X+Y}{\sim}}([e]) = [t_{X+Y}(e)]^{*}
\]

So far, the construction matches the construction of the pushout in $\catname{Hyp_{\Sigma}}$ which can be found in~\cite{bonchi_string_2022-1} and hence $s_{\sfrac{X+Y}{\sim}}$ and $t_{\sfrac{X+Y}{\sim}}$ are well-defined and $[-]$ is a homomorphism with respect to sources and targets and therefore $\sfrac{\iota_1}{\sim}$ and $\sfrac{\iota_{2}}{\sim}$ are.

Recall that $<$ is essentially a transitive closure of $<^{\mu}$ and hence we can first define 
    \[<_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}[u]_{E})
    \] and 
    \[
        <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]_{V})
    \] where $\iota_*$ are injections into $\sfrac{V_{X} + V_{Y}}{\sim} + \sfrac{E_{X} + E_{Y}}{\sim}$.
    We will consider several cases.
    First, assume that $u$ is an element of $V_{X+Y}$ and
    \begin{enumerate}
        % \item $u$ has a pre-image $z$ in $V_{Z}$, $[z) \not = \varnothing$.
        %       The last requirement means that there necessarily exist $x = <_{X}^{\mu}(f_{E}(z)) = f_{E}(<_{Z}^{\mu}(z))$ and $y = <_{Y}^{\mu}(g_{E}(z)) = g_{E}(<_{Z}^{\mu}(z))$.
        %       We then define $<_{\sfrac{X+Y}{\sim}}^{\mu}([u]) = [<_{X+Y}^{\mu}(u)]$
        \item $u$ has a pre-image $z$ in $V_{Z}$ (by discreteness, $[z) = \varnothing$) and $[f_{V}(z)) = \varnothing$, $[g_{V}(z)) \not = \varnothing$.
              The last condition means that there exists $e_y = <_{Y}^{\mu}(\iota_{V_{Y}}(g_{V}(z)))$.
              We then define 
              \[
                <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]) = [g_{V};\iota_{V_{Y}};<_{Y}^{\mu};\iota_{2,E}(z)] = [\iota_{2,E}(e_y)]
              \]
        \item $u$ has a pre-image $z$ in $V_{Z}$, (by discreteness, $[z) = \varnothing$) and $[f_{V}(z)) \not = \varnothing$ $[g_{V}(z)) = \varnothing$.
              The one before the last condition means that there exists $e_x = <_{X}^{\mu}(\iota_{V_{X}}(f_{V}(z)))$.
              We then define 
              \[
                <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]) = [f_{V};\iota_{V_{X}};<_{X}^{\mu};\iota_{1,E}(z)] = [\iota_{1,E}(e_x)]
              \]
        \item $u$ has no pre-image in $V_{Z}$ and $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(u)) = e'$
        \[
            <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]) = [<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(u))] = [e']
        \]
        \item $u$ has no pre-image in $V_{Z}$ and $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(u))$ is undefined (\textit{i.e.}, $[u) = \varnothing$). 
              If there exists $[v]$ such that $<_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v])) = [e']$ and such that there is an \textit{undirected} path from $[u]$ to $[v]$, then we define
              \[ 
                <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]) = <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v])
              \]
    \end{enumerate}
    % \question{In the case when images of $f$ and $g$ are top-level the function is undefined. Shall we introduce bottom?}
    Now, assume that $u$ is an element of $E_{X+Y}$.
    We have the only case of $u$ having no pre-image in $E_{Z}$ because it is empty.
    Consider two cases.
    \begin{enumerate}
    \item  $<_{X+Y}^{\mu}(\iota_{E_{X} + E_{Y}}(u)) = u_1$.
            Then we define
        \[
            <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}[u]) = [<_{X+Y}^{\mu}(\iota_{E_{X} + E_{Y}}(u))]
        \]
    \item $[u) = \varnothing$ and there exists $[v]$ such that $<_{\sfrac{X+Y}{\sim}}^{\mu}([v]) = [e']$ and such that there is an undirected path from $[u]$ to $[v]$, we define 
        \[
            <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}[u]) = <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v])
        \] 
    \end{enumerate}
    % We will also build this relation in steps, first we do steps for vertices (1) - (3), then step (1) for edges and finally, we will interleave step (2) for edges and step (4) for vertices until fixed point.
    % Essentially, this interleaving means that vertices inherit the relation from incident edges and vice versa.
    Clearly, all the cases above are disjoint.

    Let's show that the definition does not depend on a particular choice of $[v]$ in (4).
    Suppose, $[u) = \varnothing$ and there exist $[v_1]$ and $[v_2]$ such that $<_{\sfrac{X+Y}{\sim}}^{\mu}([v_1]) = [e_1]$ and $<_{\sfrac{X+Y}{\sim}}^{\mu}([v_2]) = [e_2]$ and $[e_1] \not = [e_2]$ and such that there is a path from $[u]$ to $[v_1]$ and from $[u]$ to $[v_2]$.
    First, both $v_1$ and $v_2$ should have a pre-image in $V_{Z}$, otherwise it would mean that there is a path from, \textit{e.g.},  $u$ to $v_1$ and that $[v_1) \not = \varnothing$ which would imply that $[u) \not = \varnothing$ and we get contradiction.
    Then consider several cases.
    \begin{itemize}
        \item $v_1 = f_{V};\iota_{1,V}(z_1)$ and $v_2 = f_{V};\iota_{1,V}(z_2)$.
        \begin{itemize}
            \item $u$ has a pre-image in $V_{X}$. Then $[f_{V}(z_1)) = \varnothing$ and $[g_{V}(z_1)) \not = \varnothing$, similarly for $z_2$.
                  Hence, 
                  \[
                    <_{\sfrac{X+Y}{\sim}}^{\mu}([v_1]) = [g_{V};\iota_{V_{Y}};<_{Y}^{\mu};\iota_{2,E}(z_1)] = [e_1]
                  \]
                  and
                  \[
                    <_{\sfrac{X+Y}{\sim}}^{\mu}([v_2]) = [g_{V};\iota_{V_{Y}};<_{Y}^{\mu};\iota_{2,E}(z_2)] = [e_2]
                  \]
                  Then, $[e_1] \not = [e_2]$ implies
                  \[
                    [g_{V};\iota_{V_{Y}};<_{Y}^{\mu};\iota_{2,E}(z_1)] \not = [g_{V};\iota_{V_{Y}};<_{Y}^{\mu};\iota_{2,E}(z_2)]
                  \] and because $[]_{E}$ and $\iota_{2,E}$ are injective, it further means that
                  \[
                    <_{Y}^{\mu}(g_{V};\iota_{V_{Y}}(z_1)) \not = <_{Y}^{\mu}(g_{V};\iota_{V_{Y}}(z_2))
                  \]
                  which contradicts assumption (3) from~\ref{pushout:assumptions}.
                  \item Case when $u$ has a pre-image in $V_{Y}$ is symmetric.
                \end{itemize}
                \item $v_1 = f_{V};\iota_{1,V}(z_1)$ and $v_2 = g_{V};\iota_{1,V}(z_2)$
                \begin{itemize}
                    \item $u$ has a pre-image in $V_{X}$. Then, $[f_{V}(z_1)) = \varnothing$, $[g_{V}(z_1)) \not = \varnothing$, $[f_{V}(z_2)) = \varnothing$, $[g_{V}(z_2)) \not = \varnothing$.
                          This contradicts the assumption (4) from~\ref{pushout:assumptions}.
                    \item Case when $u$ has a pre-image in $V_{Y}$ is symmetric.
                \end{itemize}
                \item The remaining two cases are symmetric.
    \end{itemize}

    Let's check that this is a well-defined function.
    Suppose $t_1, t_2$ are vertices and $t_1 \sim t_2$, then $<_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X}+V_{Y}}{\sim}}[t_1]) = <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[t_2])$.
    $t_1 \sim t_2$ implies there is $z$ such that $t_1 = f_{V};\iota_{1,V}(z)$ and $t_2 = g_{V};\iota_{2,V}(z)$ and two cases
    \begin{itemize}
        \item $[f_{V}(z)) \not = \varnothing$ and $[g_{V}(z)) = \varnothing$
        \item $[f_{V}(z)) = \varnothing$ and $[g_{V}(z)) \not = \varnothing$
    \end{itemize} 
    Consider the first case. Then, by definition.
    \[
        <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[t_1]) = [f_{V};\iota_{V_{X}};<_{X}^{\mu};\iota_{1,E}(z)]
    \]
    and
    \[
        <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[t_2]) = [f_{V};\iota_{V_{X}};<_{X}^{\mu};\iota_{1,E}(z)]
    \]
    so we have well-defined-ness on the nose. 
    We do not need to check well-defined-ness for edges as $e_1 \sim e_2$ just implies $e_1 = e_2$ for edges.

>>>>

\begin{theorem}[Existence of pushouts in $\catname{EHyp_{\Sigma}}$]
\label{th:existence_of_pushouts}
Consider the following span in $\catname{EHyp_{\Sigma}}$
% https://q.uiver.app/#q=WzAsMyxbMCwwLCJaIl0sWzIsMCwiWCJdLFswLDIsIlkiXSxbMCwxLCJmIl0sWzAsMiwiZyIsMl1d
\[\begin{tikzcd}
	Z && X \\
	\\
	Y
	\arrow["f", from=1-1, to=1-3]
	\arrow["g"', from=1-1, to=3-1]
\end{tikzcd}\]
such that
\begin{enumerate}
\label{pushout:assumptions}
    \item $Z$ is a \textit{discrete} e-hypergraph.
    \item $f,g$ are monos.
    \item \label{assumption:equal_predecessors} $[f_{V}(v_i)) = [f_{V}(v_j))$ and $[g_{V}(v_i)) = [g_{V}(v_j))$ for all $v_{i},v_{j}$ in $V_{Z}$.
    \item \label{assumption:non_ambiguous_predecessors} If $[f_{V}(v)) \not = \varnothing$ then $[g_{V}(v)) = \varnothing$ and if $[g_{V}(v)) \not = \varnothing$ then $[f_{V}(v)) = \varnothing$.
    \item $\consistency(f_{V}(v_i)) = \consistency(f_{V}(v_j))$ and $\consistency(g_{V}(v_i)) = \consistency(g_{V}(v_j))$ for all $v_i,v_j$ in $V_{Z}$.
\end{enumerate}    
then the pushout $X +_{f,g} Y$ exists.
\end{theorem}
\begin{proof}
    Consider the diagram below.
    \[
        \adjustbox{scale=1.25}{
            \begin{tikzcd}
            Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
            Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                            &                                                                  & Q
            \end{tikzcd}}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, <_{X+Y} \}
    \]
    where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, <_{X+Y}, \consistency_{X+Y}$.
    We will omit labels as they are irrelevant to pushout construction.
    Consider relations
    \begin{align*}    
    S_{V} &= \{\\
          &\;(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}\\
          &\cup\\
          &\{\\
          &\;(y_j,x_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}\\
          &\cup\\
          &\{\\
          &\;(x,x)\;\text{where}\; x \in V_{X} + V_{Y}\\
          &\}\\
        \\
    S_{E} &= \{(x,x) \text{ where } x \in E_{X} + E_{Y}\}
    \end{align*}
    % \begin{align*}
    % S_{E} &= \{\\
    % &(x_i,y_j) \in (E_{X} + E_{Y}) \times (E_{X} + E_{Y})\; |\\
    % &\exists z \in E_{Z} \; . \; x_i = f_{E};\iota_{1,E}(z) \text{ and } y_j = g_{E};\iota_{2,E}(z)\\
    % &\text{ where $x_i \in E_{X}$ and $y_j \in E_{Y}$}\\
    % &\}
    % \end{align*}
and let relations $R_{V},R_{E}$ be their transitive closures respectively.

We then quotient the set vertices and edges in $X + Y$ by $R_{V}$ and $R_{E}$ respectively.
    \begin{align*}
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = \{&\\
            &\sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})}, \sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})}, \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})}, \consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}}, <_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\\
        & \}    
    \end{align*}
We will refer to $\sim (R_{V},R_{E})$ just as $\sim$, \textit{e.g.}, by writing $\sfrac{V_{X} + V_{Y}}{\sim}$ and concrete relation will be clear from the context.
Where necessary we will refer to $S_{V}$ and $S_{R}$ as $\sim_{S}$.
We have 
\[
    \sfrac{s_{X+Y}}{\sim} : \sfrac{E_{X} + E_{Y}}{\sim} \to (\sfrac{V_{X} + V_{Y}}{\sim})^{*}
\]
and there is an obvious surjective function $[-]_{V} : (V_{X} + V_{Y}) \to (\sfrac{V_{X} + V_{Y}}{\sim})$ that maps elements to their equivalence classes and $[-]_{V}^{*}$ is its extension to sequences.
There is also $[-] : E_{X} + E_{Y} \to \sfrac{E_{X} + E_{Y}}{\sim}$ and we will omit subscripts as the correct type will be clear from the argument.
We then define 
\[
    \sfrac{s_{X+Y}}{\sim}([e]) = s_{X+Y};[-]^{*}(e) = [s_{X+Y}(e)]^{*}
\]
We will also use subscripts when it is important to tell if an element of $E_{X} + E_{Y}$ has a pre-image in either $E_{X}$ or $E_{Y}$ by writing $e_{x}$ or $e_{y}$.
Similarly, we will write $v_{x}$ to refer to a vertex with a pre-image in $V_{X}$.
Likewise, 
\[
    t_{\sfrac{X+Y}{\sim}}([e]) = [t_{X+Y}(e)]^{*}
\]

These definitions make $([]_{V},[]_{E})$ automatically a homomorphism with respect to source and target maps.
These maps are also automatically well-defined functions since $[e_1] = [e_2]$ implies $e_1 = e_2$ since the mappings for edges are mono.

% So far, the construction matches the construction of the pushout in $\catname{Hyp_{\Sigma}}$ which can be found in~\cite{bonchi_string_2022-1} and hence $s_{\sfrac{X+Y}{\sim}}$ and $t_{\sfrac{X+Y}{\sim}}$ are well-defined and $[-]$ is a homomorphism with respect to sources and targets and therefore $\sfrac{\iota_1}{\sim}$ and $\sfrac{\iota_{2}}{\sim}$ are.

Recall that $<$ is essentially a transitive closure of $<^{\mu}$ and hence we can first define 
    \[<_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}[e])
    \] and 
    \[
        <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[u])
    \] where $\iota_*$ are injections into $\sfrac{V_{X} + V_{Y}}{\sim} + \sfrac{E_{X} + E_{Y}}{\sim}$.
    We will consider several cases.
    First, assume that $u$ is an element of $V_{X+Y}$ and
    \begin{enumerate}
        \item If there exists $v$ such that $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(v))$ is defined and $u \sim v$
              we let
              \[
                <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([u])) = [<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(v))]
              \]
        \item $u$ has no pre-image in $V_{Z}$ and $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(u))$ is undefined (\textit{i.e.}, $[u) = \varnothing$).
              If there exists $v$ such that $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(v)) = e'$ and such that there is an \textit{undirected} path from $[u]$ to $[v]$, then we define
              \[ 
                <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]) = <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v])
              \]
    \end{enumerate}
    There is no case when $u$ has a pre-image in $V_{Z}$ and yet $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(u))$ is undefined and there exists $v$ such that $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(v)) = e$ and there is a path from $[u]$ to $[v]$.
    The existence of a path would mean that there is $u' \sim u$ and $v' \sim v$ such that there is a path from $u'$ to $v'$ and since $u$ has a pre-image in $V_{Z}$, $u = f_{V};\iota_{1,V}(z_1)$.
    $v' \sim v$ means that both $v'$ and $v$ have a pre-image in $V_{Z}$ (unless $v' = v$ in which case $[u') \not = \varnothing$ as there is a path from $u'$ to $v'$) and since $u = f_{V};\iota_{1,V}(z_1)$ and $[u) = \varnothing$ $f_{V};\iota_{1,V}(z) = \varnothing$ for all $z$.
    Since $[v) \not = \varnothing$ and $v' \sim v$, $v$ is necessarily in the image of $g_{V};\iota_{2,V}$ and for all $z$ $[g_{V};\iota_{2,V}(z)) \not = \varnothing$ as well as for some $z'$ such that $u'' = g_{V};\iota_{2,V}(z')$ and $u'' \sim_{S} u$ which would mean that the case (1) is applicable to $u$.
    Similarly, when $u = g_{V};\iota_{2,V}(z)$.

    Otherwise, we leave $<_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{V_{X} + V_{Y}}{\sim}([u]))$ undefined.
    % \question{In the case when images of $f$ and $g$ are top-level the function is undefined. Shall we introduce bottom?}
    Now, assume that $e$ is an element of $E_{X+Y}$.
    Consider two cases.
    \begin{enumerate}
    \item  $<_{X+Y}^{\mu}(\iota_{E_{X} + E_{Y}}(e)) = e'$.
            Then we define
        \[
            <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}[e]) = [<_{X+Y}^{\mu}(\iota_{E_{X} + E_{Y}}(e))]
        \]
    \item $[e) = \varnothing$ and there exists $v$ such that $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(v)) = e'$ and such that there is an undirected path from $[e]$ to $[v]$, we define 
        \[
            <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}[e]) = <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v])
        \] 
    \end{enumerate}
    Otherwise we leave $<_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{E_{X} + E_{Y}}{\sim}([e]))$ undefined.
    % We will also build this relation in steps, first we do steps for vertices (1) - (3), then step (1) for edges and finally, we will interleave step (2) for edges and step (4) for vertices until fixed point.
    % Essentially, this interleaving means that vertices inherit the relation from incident edges and vice versa.
    Clearly, all the cases above are disjoint.

    Let's show that the definition does not depend on a particular choice of $v$ in (1).
    Suppose there exist $v_1$ and $v_2$ such that $v_1 \not = v_2$ and $[v_1) \not = \varnothing$ and $[v_2) \not = \varnothing$ and such that $u \sim v_1$ and $u \sim v_2$.
    Then, it must be the case that $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(v_1)) = <_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(v_2))$.
    Consider cases.
    \begin{itemize}
        \item $u = v_1$ and $u \sim_{S} v_2$. Then there exists $z \in V_{Z}$ such that $u = f_{V};\iota_{1,V}(z) = v_1$ and $v_2 = g_{V};\iota_{2,V}(z)$.
              Since $[v_1) \not = \varnothing$, it must be the case that $[v_2) = \varnothing$ according to assumption \ref{assumption:non_ambiguous_predecessors} and we get a contraction to $[v_2) \not = \varnothing$.
        \item $u = v_1$ and $u \sim v_2$. The latter implies that $u$ has a pre-image in $V_{Z}$ and so does $v_2$.
              For both $[v_1) \not = \varnothing$ and $[v_2) \not = \varnothing$ they should be both in the image of $f_{V};\iota_{1,V}$ or $g_{V};\iota_{2,V}$ and according to assumption \ref{assumption:equal_predecessors} their sets of predecessors must be equal.
        \item $u \sim_{S} v_1$ and $u \sim_{S} v_2$. This implies $u = f_{V};\iota_{1,V}(z_1) = f_{V};\iota_{1,V}(z_2)$ and $v_{1} = g_{V};\iota_{2,V}(z_1)$ and $v_{2} = g_{V};\iota_{2,V}(z_2)$.
              By assumption \ref{assumption:equal_predecessors} it must be the case that $[g_{V}(z_1)) = [g_{V}(z_2))$ which implies that $[v_1) = [v_2)$.
              The case when $u = g_{V};\iota_{2,V}(z_1)$ is symmetric.
        \item $u \sim v_1$ and $u \sim v_2$. Same as above, by assumption \ref{assumption:non_ambiguous_predecessors} both $v_1$ and $v_2$ should be in the image of $f_{V};\iota_{1,V}$ or $g_{V};\iota_{2,V}$ and by assumption \ref{assumption:equal_predecessors} $[v_1) = [v_2)$.
    \end{itemize}
